---
title: "SBC Chinook Catch and Release Summary"
subtitle: "2011-2023"
output:
  pdf_document:
    extra_dependencies: ["float"]
  html_notebook: default
editor_options:
  chunk_output_type: console
---

## Data Sources

-   2010-2021 effort, catch and released numbers are based on DFO provided data
-   2023 versus 5-yr recent average are based on the South Coast Assessment Creel Survey Summary Report from September 2023

## Notes

-   sub-legal = smaller than the legal size limit
-   encounters = total number of kept, released-legal, and released sub-legal chinook

```{r load library, echo=FALSE,message=FALSE,warning=FALSE}

knitr::opts_chunk$set(fig.pos = "H", out.extra = "")

options(scipen=1000)
library(data.table)
library(tidyverse)
library(RColorBrewer)
library(ggpubr)
library(stringr)
library(patchwork)
library(ggplot2)
library(dplyr)
require(readxl)
```

```{r data read,  echo=FALSE,message=FALSE,warning=FALSE}
data<-read_excel("data/catch and release data april 2023 version.xlsx",sheet="chinook catch and release")

effort<-read_excel("data/catch and release data april 2023 version.xlsx",sheet="effort by target species")
```

```{r data read new 2023 effort data,  echo=FALSE,message=FALSE,warning=FALSE}
new.effort.data<-fread("data/South_Coast_Recreational_effort_2012-2023.csv")
#names(new.effort.data)

new.effort.area.merge<-fread(("data/pfma to area groupings merge.csv"))

e22.prep<-new.effort.data%>%
  select(PFMA,YEAR,MONTH,MANAGEMENT,ESTIMATE,ESTIMATE,STANDARD_ERROR,PERCENT_STANDARD_ERROR)

e22<-merge(e22.prep,new.effort.area.merge,by="PFMA")%>%
  mutate(PFMA=factor(PFMA,levels=c("PFMA 11","PFMA 111","PFMA 12",
                                   "PFMA 13","PFMA 14","PFMA 15",
                                   "PFMA 16","PFMA 17","PFMA 18",
                                   "PFMA 19","PFMA 20","PFMA 21",
                                   "PFMA 22","PFMA 23","PFMA 24","PFMA 25","PFMA 26","PFMA 27",
                                   "PFMA 121","PFMA 123","PFMA 124","PFMA 125","PFMA 126","PFMA 127",
                                   "PFMA 28","PFMA 29")))


#x<-data.frame(PFMA=unique(e22.prep$PFMA))
#write.csv(x,"data/pfma to area groupings merge.csv")

years=seq(2012,2023,1)

e22$months=factor(e22$MONTH,levels=c("January","February","March","April","May","June","July","August","September","October","November","December"))

```

```{r 2023 data read and prep,  echo=FALSE,message=FALSE,warning=FALSE}
d22<-read_excel("data/catch and release data april 2023 version.xlsx",sheet="2023 sept creel survey totals")

d22.long<-d22%>%mutate(encounters=kept+released)%>%
pivot_longer(c(4:7),names_to="disposition",values_to="estimate")%>%
  mutate(area2=factor(area))
```

```{r 2023 per SE data read,  echo=FALSE,message=FALSE,warning=FALSE}
perse<-read_excel("data/catch and release data april 2023 version.xlsx",sheet="percent SE 2023 creel summary")%>%mutate(area2=factor(area))

```

```{r data prep, echo=FALSE,message=FALSE,warning=FALSE}

gg.prep<-data%>%
  select(PFMA,CREEL_SUB_AREA,YEAR,MONTH,DISPOSITION,ADIPOSE_MARK,ESTIMATE,STANDARD_ERROR,PERCENT_STANDARD_ERROR,AREA=`Area Grouping`)

years=seq(2011,2023,1)

area.region.merge=data.frame(AREA=c("Alberni (23A,B)","Barkley (23C-Q)","Clayquot","Kyuquot",
                                    "Nitinat","Nootka/Esp","Quatsino",
                                    "NWVIoffshore(125,126,127)","SWVIoffshore(121,123,124)",
                                    "JDFeast(19cde,20cd)","JDFwest(Renfrew 20abe)",
                                    "nGST(13,14,15,16)","nJST(12)","sGST(17,18,19ab,28,29)","upperQCS(11,111)"),
                             region=c("WCVI","WCVI","WCVI","WCVI","WCVI","WCVI","WCVI","WCVI","WCVI",
                                      "JDF","JDF",
                                      "SOG/ECVI","SOG/ECVI","SOG/ECVI","SOG/ECVI"))

months=factor(c("May","June","July","August","September"),levels=c("May","June","July","August","September"))
month.merge<-data.frame(MONTH=c(5,6,7,8,9),Month=months)

gg.effort.prep<-merge(effort,month.merge,by="MONTH")

```

\newpage

# EFFORT DATA INCLUDING 2023

-   data from DFO (2002-2021 from DFO database, 2022 and 2023 digitized from SC creel reports)
-   2002-2023 by month by year by PFMA by management (ISBM versus AABM)
-   effort is total boat trips
-   Note that this is all effort, not Chinook directed effort
-   2023 marginally less effort than 2022, continues lower effort since 2018

```{r total effort all sbc new data,echo=FALSE,message=FALSE,warning=FALSE,fig.height=4,fig.width=5,fig.cap="Total effort in number of boat trips for all SBC areas."}
e22.tot<-e22%>%group_by(YEAR)%>%summarise(total=sum(ESTIMATE,na.rm=TRUE))

ggplot(e22.tot,aes(x=YEAR,y=total))+
  geom_col()+
  theme_bw()+
  labs(x="Year",y="Total Effort SBC (boat trips)")
```

\newpage

-   effort by PFMA for 2002-2023
-   less effort in most areas in 2023 versus 2022
-   some areas (mostly in WCVI) increased effort
-   effort in Fraser approach (28/29) increased slightly from 2023

```{r total effort all sbc by area by year, echo=FALSE,message=FALSE,warning=FALSE,fig.height=7.5,fig.width=7,fig.cap="Total effort in number of boat trips by area for all SBC areas."}

e22.byarea<-e22%>%group_by(PFMA,YEAR,AREA)%>%summarise(effort=sum(ESTIMATE,na.rm=TRUE))%>%
  mutate(PFMA=factor(PFMA,levels=c("PFMA 11","PFMA 111","PFMA 12",
                                   "PFMA 13","PFMA 14","PFMA 15",
                                   "PFMA 16","PFMA 17","PFMA 18",
                                   "PFMA 19","PFMA 20","PFMA 21",
                                   "PFMA 22","PFMA 23","PFMA 24","PFMA 25","PFMA 26","PFMA 27",
                                   "PFMA 121","PFMA 123","PFMA 124","PFMA 125","PFMA 126","PFMA 127",
                                   "PFMA 28","PFMA 29")))

cols<-brewer.pal(7,"Set1")[c(1:5,7)]

ggplot(e22.byarea,aes(x=YEAR,y=effort,color=AREA))+
  geom_point()+
  geom_line()+
  theme_bw()+
  scale_color_manual(values=cols)+
  facet_wrap(~PFMA,scale="free_y",ncol=3)+
  expand_limits(y=0)+
  theme(legend.position="bottom")+
  labs(x="Year",y="Effort (boat trips)",color="Area")

```

\newpage

```{r Kept and released by PFMA and Region, echo=FALSE,message=FALSE,warning=FALSE,fig.height=7.8,fig.width=7,fig.cap="Chinook kept and released by Region and PFMA"}

pfma.lkp <- read.csv("data/pfma_region_lookup.csv") %>%
  mutate(PFMA = paste("PFMA", PFMA))

# make new df for this plot
gg.rel <- gg.prep

# Combine legal, sub-legal, and unspecified to single "Released" 
gg.rel$DISPOSITION[gg.rel$DISPOSITION == "Released Legal"] <- "Released"
gg.rel$DISPOSITION[gg.rel$DISPOSITION == "Released Sub-legal"] <- "Released"
gg.rel$DISPOSITION[gg.rel$DISPOSITION == "Released Unspecified"] <- "Released"

gg.rels <- gg.rel %>%
  merge(pfma.lkp, by = "PFMA") %>%
  group_by(REGION, YEAR, DISPOSITION) %>%
  summarise(total = sum(ESTIMATE, na.rm = TRUE)) %>%
  merge(pfma.lkp, by = "REGION")

r1 <- ggplot(gg.rels, aes(x = YEAR, y = total, color = DISPOSITION)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~REGION) +
  labs(x = "Year", y = "Number of Chinook") +
  theme_bw() + 
  scale_color_brewer(palette = "Set1")

r2 <- ggplot(gg.rels, aes(x = YEAR, y = total, color = DISPOSITION)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~PFMA) +
  labs(x = "Year", y = "Number of Chinook") +
  theme_bw() + 
  scale_color_brewer(palette = "Set1")

ggarrange(r2, r1, nrow = 2)


```

\newpage

```{r total effort all sbc by area by year only may-sept, echo=FALSE,include=FALSE,message=FALSE,warning=FALSE,fig.height=8.5,fig.width=7,fig.cap="Total effort in number of boat trips by area for all SBC areas-ONLY MAY to SEPTEMBER."}

e22.byareaonlysomemonths<-e22%>%filter(MONTH%in%c("May","June","July","August","September"))%>%
  group_by(PFMA,YEAR,AREA)%>%
  summarise(effort=sum(ESTIMATE,na.rm=TRUE))

cols<-brewer.pal(7,"Set1")[c(1:5,7)]

ggplot(e22.byarea,aes(x=YEAR,y=effort,color=AREA))+
  geom_point()+
  geom_line()+
  theme_bw()+
  scale_color_manual(values=cols)+
  facet_wrap(~PFMA,scale="free_y",ncol=3)+
  theme(legend.position="bottom")+
  labs(x="Year",y="Effort (boat trips)",color="Area")

```

\newpage

```{r total by disposition all sbc,echo=FALSE,message=FALSE,warning=FALSE,fig.height=3,fig.width=5,fig.cap="Total encounters (column height) by disposition (kept, released and released sub-legal) Chinook for all SBC areas."}
gg.totalbydis<-gg.prep%>%
  group_by(YEAR,DISPOSITION)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))


c1<-ggplot(gg.totalbydis,aes(x=YEAR,y=total,fill=DISPOSITION))+
  geom_col()+
  scale_fill_brewer(palette="Set1")+
  scale_x_continuous(labels=as.character(years),breaks=years)+
  theme_bw()+
  labs(x="Year",y="Chinook",fill="")+
  geom_text(aes(label = total), position=position_stack(vjust=0.5), colour = "black",size=2)+
  theme(legend.position="bottom")
```

```{r total effort all sbc, echo=FALSE,message=FALSE,warning=FALSE,fig.height=3,fig.width=5,fig.cap="Total Chinook directed effort by year and month for all SBC areas."}
gg.efforttotal<-gg.effort.prep%>%filter(TARGET=="Chinook")%>%
  group_by(YEAR,Month)%>%
  summarise(total=sum(TRIPS,na.rm=TRUE))

e1<-ggplot(gg.efforttotal,aes(x=YEAR,y=total,fill=factor(Month)))+
  geom_col()+
  theme_bw()+
  scale_x_continuous(labels=as.character(seq(2014,2020,1)),breaks=seq(2014,2020,1))+
  scale_fill_brewer(palette="Set1")+
  labs(x="Year",y="Total Chinook Effort (trips)",fill="")+
  geom_text(aes(label = total), position=position_stack(vjust=0.5), colour = "black",size=3)+
  theme(legend.position="bottom")

#ggsave("figures/chinook total by disposition all SBC.png",height=5,width=6,dpi=600)
```

# ENCOUNTERS

-   total encounters lower in 2023 than in 2022
-   we don't have sub-legal/legal breakdown for 2022 and 2023 yet
-   total encounters in 2023 still higher than pre-2021 years
-   total encounters in 2023 \~ 430,000 Chinook (172k kept and 268k released)

```{r total catch release effort all sbc2, echo=FALSE,message=FALSE,warning=FALSE,fig.height=5,fig.width=6.5,fig.cap="Top: Total encounters (column height) by disposition (kept, released and released sub-legal) Chinook for all SBC areas. "}

# Removed bottom plot because we weren't comparing apples to apples with effort
#ggarrange(c1,e1,ncol=1)
c1

#ggsave("figures/chinook directed effort by year and disposition summary.png",height=6,width=7,dpi=600)
```

```{r total catch release effort all sbc,include=FALSE, echo=FALSE,message=FALSE,warning=FALSE,fig.height=8,fig.width=5,fig.cap="Top: Total encounters (column height) by disposition (kept, released and released sub-legal) Chinook for all SBC areas. Bottom: Total Chinook directed effort by year and month for all SBC areas."}
# Muted for now

gg.effortbyarea<-gg.effort.prep%>%filter(TARGET=="Chinook")%>%
  group_by(LRG_AREA_NME,YEAR)%>%
  summarise(total=sum(TRIPS,na.rm=TRUE))

ggplot(gg.effortbyarea,aes(x=YEAR,y=total))+
  geom_point()+geom_line()+
  theme_bw()+
  facet_grid(LRG_AREA_NME~.,switch="y",scale="free_y")+
  theme(strip.text.y.left=element_text(angle=0,size=8))+
  labs(x="Year",y="Total Chinook Directed Effort")+
  expand_limits(y=0)

#ggsave("figures/chinook directed effort by area.png",height=6,width=7,dpi=600)

```

\newpage

# KEPT AND RELEASED BY AREA

-   Kept and released various changes by area (down in some areas, up in others)

```{r total across disposition by area,echo=FALSE,message=FALSE,warning=FALSE,fig.height=7,fig.width=8,fig.cap="Kept and released Chinook by area."}
gg.totalbydisbyarea<-gg.prep%>%
  group_by(AREA,YEAR,DISPOSITION)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))

# Combine legal, sub-legal, and unspecified to single "Released" 
gg.totalbydisbyarea$DISPOSITION[gg.totalbydisbyarea$DISPOSITION == "Released Legal"] <- "Released"
gg.totalbydisbyarea$DISPOSITION[gg.totalbydisbyarea$DISPOSITION == "Released Sub-legal"] <- "Released"
gg.totalbydisbyarea$DISPOSITION[gg.totalbydisbyarea$DISPOSITION == "Released Unspecified"] <- "Released"

# Filter out areas that are broken down in detail in next figure
gg.totalbydisbyarea.cut <- gg.totalbydisbyarea %>%
  filter(AREA != "JDFeast(19cde,20cd)") %>%
  filter(AREA != "JDFwest(Renfrew 20abe)") %>%
  filter(AREA != "nGST(13,14,15,16)") %>%
  filter(AREA != "sGST(17,18,19ab,28,29)")

ggplot(gg.totalbydisbyarea.cut,aes(x=YEAR,y=total,fill=DISPOSITION))+
  geom_col()+
  scale_fill_brewer(palette="Set1")+
  scale_x_continuous(labels=as.character(years),breaks=years)+
  theme_bw()+
  facet_grid(AREA~DISPOSITION,scale="free_y",switch="y")+
  labs(x="Year",y="Chinook",fill="")+
 #geom_text(aes(label = total), position=position_stack(vjust=1), colour = "black",size=1,angle=90)+
  theme(legend.position="bottom",strip.text.y.left=element_text(angle=0,size=6)) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

\newpage

-   in key MSF areas, we don't have the sub-area level catch and release numbers to divide these up, so they are only rough estimates except for nGST

```{r total across disposition by area for just some areas stacked, echo=FALSE,message=FALSE,warning=FALSE,fig.height=8,fig.width=8,fig.cap="Kept, released and released sub-legal Chinook by area for proposed MSF areas in SOG and JDF."}
gg.totalbydisbyarea.msf<-gg.prep%>%
  filter(AREA%in%c("JDFeast(19cde,20cd)","JDFwest(Renfrew 20abe)","sGST(17,18,19ab,28,29)","nGST(13,14,15,16)"))%>%
  group_by(AREA,YEAR,DISPOSITION)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))

ggplot(gg.totalbydisbyarea.msf,aes(x=YEAR,y=total,fill=DISPOSITION))+
  geom_col()+
  scale_fill_brewer(palette="Set1")+
  scale_x_continuous(labels=as.character(years),breaks=years)+
  theme_bw()+
  facet_grid(AREA~.,scale="free_y",switch="y")+
  labs(x="Year",y="Number of Chinook",fill="")+
  geom_text(aes(label = total), position=position_stack(vjust=.5), colour = "black",size=2,angle=0)+
  theme(legend.position="bottom",strip.text.y.left=element_text(angle=0,size=6))

#ggsave("figures/chinook total by disposition by area msf filter.png",height=8,width=8,dpi=600)
```

\newpage

# CHINOOK DIRECTED EFFORT

-   Chinook directed effort from 2014-2020 by area and month
-   We don't have this for 2021-2023, so can't add those years

```{r effort by area month,echo=FALSE,message=FALSE,warning=FALSE,fig.height=7,fig.width=8,fig.cap="Chinook directed effort by area and month (2014-2020)"}

gg.effortbyarea<-gg.effort.prep%>%
  group_by(area=LRG_AREA_NME,YEAR,Month)%>%
  summarise(effort=sum(TRIPS,na.rm=TRUE))

# Revert to old version until full data becomes available
gg.effortbyarea.cut <- gg.effortbyarea %>%
  filter(!is.na(area))

ggplot(gg.effortbyarea.cut,aes(x=YEAR,y=effort,fill=Month))+
  #geom_line()+geom_point()+
  geom_col(position=position_dodge(preserve="single"))+
  scale_fill_brewer(palette="Set1")+
  theme_bw()+
  theme(legend.position = "bottom")+
  facet_wrap(~area,ncol=2)+
  labs(x="Year",y="Effort (trips)",color="")

```

\newpage

```{r total by disposition stacked by area,include=FALSE,echo=FALSE,message=FALSE,warning=FALSE,fig.height=9,fig.width=8,fig.cap="Kept, released and released sub-legal by area. Total column height is encounters."}
# Include = FALSE for now as it's redundant of figure 2
gg.totalbydisbyarea<-gg.prep%>%
  group_by(AREA,YEAR,DISPOSITION)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))

#ggplot(gg.totalbydisbyarea,aes(x=YEAR,y=total,fill=DISPOSITION))+
ggplot(gg.totalbydisbyarea,aes(x=YEAR,y=total,color=DISPOSITION))+
  #geom_col()+
  geom_point()+geom_line()+
  #scale_fill_brewer(palette="Set1")+
  scale_color_brewer(palette="Set1")+
  scale_x_continuous(labels=as.character(years),breaks=years)+
  theme_bw()+
  facet_grid(AREA~DISPOSITION,scale="free_y",switch="y")+
  labs(x="Year",y="Chinook",fill="")+
#  geom_text(aes(label = total), position=position_stack(vjust=0.5), colour = "black",size=1)+
  theme(legend.position="bottom",strip.text.y.left=element_text(angle=0),axis.text.x = element_text(angle=90))

#ggsave("figures/chinook total by disposition by area totals.png",height=10,width=8,dpi=600)
#data%>%distinct(`Area Grouping`)

```

\newpage

```{r total encounters by area,echo=FALSE,message=FALSE,warning=FALSE,include=FALSE,fig.height=9,fig.width=8,fig.cap="Total encounters by area 2011-2023. Lines fit by LOESS."}

# Include = FALSE until the area groupings are worked out

gg.encountersbyarea<-gg.prep%>%
  group_by(AREA,YEAR)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))

ggplot(gg.encountersbyarea,aes(x=YEAR,y=total))+
  geom_point()+geom_line()+
  #scale_fill_brewer(palette="Set1")+
  #scale_x_continuous(labels=as.character(years),breaks=years)+
  theme_bw()+
  geom_smooth()+
  facet_wrap(~AREA,scale="free_y",ncol=3)+
  labs(x="Year",y="Encounters")

#ggsave("figures/chinook total by disposition by area totals.png",height=10,width=8,dpi=600)
#data%>%distinct(`Area Grouping`)
```

\newpage

```{r total enucounters by area base versus recent simple, echo=FALSE,message=FALSE,warning=FALSE,include=FALSE,fig.height=5,fig.width=8,fig.cap="Percent change in encounters between base period (2011-2018) and 2019-2023."}

# Muted until area grouping data is available

gg.encountersbaseversusrecentsimple<-gg.encountersbyarea%>%
  mutate(period=ifelse(YEAR<2019,"2011-2018","2019-2023"))%>%
  group_by(AREA,period)%>%
  summarise(u=mean(total))%>%
  pivot_wider(names_from="period",values_from="u")%>%
  mutate(pchange=`2019-2023`/`2011-2018`-1)%>%
  select(AREA,pchange)%>%
  merge(area.region.merge)%>%
  mutate(factor(region,levels=c("WCVI","SOG/ECVI","JDF")))

ggplot(gg.encountersbaseversusrecentsimple,aes(x=fct_reorder(AREA,region),y=pchange,fill=region))+
  geom_col()+
  theme_bw()+
  #facet_grid(. ~ region, space="free",scale="free")+ 
  labs(y="Percent Chnage in Encounters",x="Period")+
  scale_fill_brewer(palette="Set1")+
  theme(axis.text.x=element_text(angle=90))

```

\newpage

```{r total enucounters by area base versus recent,echo=FALSE,message=FALSE,warning=FALSE,include=FALSE,fig.height=9,fig.width=8,fig.cap="Boxplot of total encounters in base period (2011-2018) versus recent years (2019-2021)."}

# Muted until area groupings are available

gg.encountersbaseversusrecent<-gg.encountersbyarea%>%
  mutate(period=ifelse(YEAR<2019,"2011-2018","2019-2021"))

ggplot(gg.encountersbaseversusrecent,aes(x=period,y=total,fill=period))+
  geom_boxplot(notch=FALSE)+
  scale_fill_brewer(palette="Set1")+
  facet_wrap(~AREA,scale="free_y",ncol=3)+
  theme_bw()+
  labs(y="Encounters",x="Period")+
  guides(fill="none")
```


