---
title: "SBC Chinook Creel Survey Summary"
subtitle: "2011-2023"
output:
  pdf_document:
    extra_dependencies: ["float"]
  html_notebook: default
editor_options:
  chunk_output_type: console
---

```{r, echo=FALSE,warning=FALSE,message=FALSE}

library(data.table)
library(tidyverse)
library(RColorBrewer)
library(ggpubr)
library(stringr)
library(patchwork)
library(ggplot2)
library(dplyr)
require(readxl)
```

```{r, echo=FALSE,warning=FALSE,message=FALSE, include=FALSE}
pfmalevels=c("11","111","12","13","14","15","16","17","18","19","20","21","121","23","123","24","124","25","125","26","126","27","127","28","29")

data<-read_excel("data/south coast creel data.xlsx",sheet="Sheet1")%>%
  mutate(pfma=factor(pfma,levels=pfmalevels))

perse<-read_excel("data/catch and release data april 2023 version.xlsx",sheet="percent SE 2023 creel summary")%>%mutate(area2=factor(area))

```


```{r,echo=FALSE,warning=FALSE,message=FALSE,include=FALSE,fig.height=6,fig.width=6,fig.cap="Creel reported effort and kept and released catch for 2022."}
data22<-data%>%filter(year==2022)

d22<-read_excel("data/catch and release data april 2023 version.xlsx",sheet="2023 sept creel survey totals")

ggplot(data22,aes(y=value,x=pfma,fill=month))+
  geom_col(position="dodge")+
  scale_fill_brewer(palette="Set1")+
  facet_wrap(metric~.,ncol=1)+
  labs("Value")+
  theme_bw()

```


\newpage

# 2023 Creel Survey Summary



```{r,echo=FALSE,warning=FALSE,message=FALSE,fig.height=6,fig.width=6,fig.cap="Creel reported effort and kept and released catch for 2023."}
data23<-data%>%filter(year==2023)

ggplot(data23,aes(y=value,x=pfma,fill=month))+
  geom_col(position="dodge")+
  scale_fill_brewer(palette="Set1")+
  facet_wrap(metric~.,ncol=1)+
  theme_bw() +
  labs(x = "PFMA", y = "Number of Chinook")

  
```

\newpage

```{r 2023 creel survey summary,echo=FALSE,message=FALSE,warning=FALSE,fig.height=7,fig.width=8,fig.cap="Top: Total encounters of Chinook by area and area grouping for 2023 year to date and the 5-yr average. Bottom: Percentage change in total encounters (2023 YTD versus 5-yr average)."}


d22.long<-d22%>%mutate(encounters=kept+released)%>%
pivot_longer(c(4:7),names_to="disposition",values_to="estimate")%>%
  mutate(area2=factor(area))

d22.effort<-d22.long%>%filter(disposition=="effort")

e22.1<-ggplot(d22.effort,aes(x=area2,y=estimate,fill=year))+
  geom_col(position="dodge")+
  scale_fill_brewer(palette="Set1")+
  theme_bw()+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Effort",fill="Period")+
  theme(legend.position="bottom")

d22.effort.pchange<-d22.effort%>%pivot_wider(names_from="year",values_from="estimate")%>%
  mutate(pchange=(`2023`-`5-yr`)/`5-yr`*100)%>%
  select(area2,region,disposition,pchange)

e22.2<-ggplot(d22.effort.pchange,aes(x=area2,y=pchange))+
  geom_col(position="dodge")+
  #scale_fill_brewer(palette="Set1")+
  theme_bw()+
  geom_hline(yintercept=0)+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Percent Change in Effort (2023 versus 5-yr average)")
```
\newpage
```{r 2023 creel survey summary output effort,include=TRUE,echo=FALSE,message=FALSE,warning=FALSE,fig.height=7,fig.width=8,fig.cap="Total effort by area and area grouping for 2023 year to date and the 5-yr average. Bottom: Percentage change in total effort (2023 YTD versus 5-yr average)."}
# include = TRUE, But does not represent chinook-directed effort
e22.1/e22.2

```

\newpage

```{r 2023 creel survey summary kept,echo=FALSE,message=FALSE,warning=FALSE,fig.height=9,fig.width=8,fig.cap=")."}
d22.kept<-d22.long%>%filter(disposition=="kept")

c22.1<-ggplot(d22.kept,aes(x=area2,y=estimate,fill=year))+
  geom_col(position="dodge")+
  scale_fill_brewer(palette="Set1")+
  theme_bw()+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Kept Chinook",fill="Period")+
  theme(legend.position="bottom")

d22.kept.pchange<-d22.kept%>%pivot_wider(names_from="year",values_from="estimate")%>%
  mutate(pchange=(`2023`-`5-yr`)/`5-yr`*100)%>%
  select(area2,region,disposition,pchange)

c22.2<-ggplot(d22.kept.pchange,aes(x=area2,y=pchange))+
  geom_col(position="dodge")+
  #scale_fill_brewer(palette="Set1")+
  theme_bw()+
  geom_hline(yintercept=0)+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Percent Change in Kept Catch\n(2023 versus 5-yr average)")
```

-   Kept Chinook in 2023 variable against 5-yr average, higher released numbers in Fraser approach and JDF, variable in SOG (higher in PFMA 18, lower in PFMA 16)

```{r 2023 creel survey summary output kept,echo=FALSE,message=FALSE,warning=FALSE,fig.height=7,fig.width=8,fig.cap="Top: Total kept catch of Chinook by area and area grouping for 2023 year to date and the 5-yr average. Bottom: Percentage change in total kept catch (2023 YTD versus 5-yr average)."}

c22.1/c22.2

```

\newpage

```{r 2023 creel survey summary release,echo=FALSE,message=FALSE,warning=FALSE,fig.height=9,fig.width=8,fig.cap=")."}
d22.rel<-d22.long%>%filter(disposition=="released")

r22.1<-ggplot(d22.rel,aes(x=area2,y=estimate,fill=year))+
  geom_col(position="dodge")+
  scale_fill_brewer(palette="Set1")+
  theme_bw()+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Released Chinook",fill="Period")+
  theme(legend.position="bottom")

d22.rel.pchange<-d22.rel%>%pivot_wider(names_from="year",values_from="estimate")%>%
  mutate(pchange=(`2023`-`5-yr`)/`5-yr`*100)%>%
  select(area2,region,disposition,pchange)

r22.2<-ggplot(d22.rel.pchange,aes(x=area2,y=pchange))+
  geom_col(position="dodge")+
  #scale_fill_brewer(palette="Set1")+
  theme_bw()+
  geom_hline(yintercept=0)+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Percent Change in Released Catch\n(2023 versus 5-yr average)")
```

-   released Chinook in 2023 variable, higher released numbers in Fraser approach and JDF, variable in SOG (much higher in PFMA 18, lower in PFMA 16)

```{r 2023 creel survey summary output released,echo=FALSE,message=FALSE,warning=FALSE,fig.height=7,fig.width=8,fig.cap="Top: Total released catch of Chinook by area and area grouping for 2023 year to date and the 5-yr average. Bottom: Percentage change in total released catch (2023 YTD versus 5-yr average)."}

r22.1/r22.2

```

\newpage

```{r 2023 creel survey summary encounters,echo=FALSE,message=FALSE,warning=FALSE,fig.height=7,fig.width=8,fig.cap=")."}
d22.enc<-d22.long%>%filter(disposition=="encounters")

enc22.1<-ggplot(d22.enc,aes(x=area2,y=estimate,fill=year))+
  geom_col(position="dodge")+
  scale_fill_brewer(palette="Set1")+
  theme_bw()+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Total Encounters of Chinook",fill="Period")+
  theme(legend.position="bottom")

d22.enc.pchange<-d22.enc%>%pivot_wider(names_from="year",values_from="estimate")%>%
  mutate(pchange=(`2023`-`5-yr`)/`5-yr`*100)%>%
  select(area2,region,disposition,pchange)

enc22.2<-ggplot(d22.enc.pchange,aes(x=area2,y=pchange))+
  geom_col(position="dodge")+
  #scale_fill_brewer(palette="Set1")+
  theme_bw()+
  geom_hline(yintercept=0)+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Percent Change in Total Encounters\n(2023 versus 5-yr average)")
```

-   Encounters of Chinook in 2023 variable, higher total encounters in Fraser approach and JDF, variable in SOG (much higher in PFMA 18, lower in PFMA 16)

```{r 2023 creel survey summary output encounters,echo=FALSE,message=FALSE,warning=FALSE,fig.height=7,fig.width=8,fig.cap="Top: Total encounters of Chinook by area and area grouping for 2023 year to date and the 5-yr average. Bottom: Percentage change in total encounters (2023 YTD versus 5-yr average)."}
enc22.1/enc22.2
```

\newpage

-   Uncertainty in the creel effort ranges from low to high depending on the area.

```{r 2023 creel survey summary per SE,echo=FALSE,message=FALSE,warning=FALSE,fig.height=7,fig.width=8,fig.cap="September 2023 creel survey SE percent values for visualising uncertainty."}
perse2<-perse%>%
  mutate(use=case_when(perse<=25~"relatively certain",
                       perse>=25&perse<=50~"use with caution",
                       perse>50~"high uncertainty"))

ggplot(perse2,aes(x=area2,y=perse,fill=factor(use,levels=c("relatively certain","use with caution","high uncertainty"))))+
         geom_col()+theme_bw()+
         facet_grid(disposition~region,space="free",switch="x",scale="free_x",labeller = labeller(region = label_wrap_gen(10)))+
  scale_fill_manual(values=c("dark green","orange","firebrick"))+
  labs(x="Area",y="Percent or Relative Standard Error",fill="")+
  theme(legend.position = "bottom")

ggsave("outputs/chinook percent standard error 2023 creel survey.png",dpi=600,height=6,width=7)

```

