---
title: "SBC Chinook Catch and Release Summary"
subtitle: "2011-2021"
format: html
editor: visual
editor_options:
  chunk_output_type: console
---

## Data Sources

-   2010-2021 effort, catch and released numbers are based on DFO provided data
-   2024 versus 5-yr recent average are based on the South Coast Assessment Creel Survey Summary Report from June 2024

## Notes

-   sub-legal = smaller than the legal size limit
-   encounters = total number of kept, released-legal, and released sub-legal chinook

```{r load library}
#| echo: false
#| message: false
#| warning: false

knitr::opts_chunk$set(fig.pos = "H", out.extra = "")

options(scipen=1000)
library(data.table)
library(tidyverse)
library(RColorBrewer)
library(ggpubr)
library(stringr)
library(patchwork)
library(ggplot2)
library(dplyr)
require(readxl)
```

```{r data read}
#| echo: false
#| message: false
#| warning: false
#| 
data<-read_excel("data/catch and release data aug 2024 version.xlsx",sheet="chinook catch and release")

effort<-read_excel("data/catch and release data aug 2024 version.xlsx",sheet="effort by target species")
```

```{r data read new 2022 effort data}
#| echo: false
#| message: false
#| warning: false 

new.effort.data<-fread("data/South_Coast_Recreational_effort_2012-2022.csv")
#names(new.effort.data)

new.effort.area.merge<-fread(("data/pfma to area groupings merge.csv"))

e22.prep<-new.effort.data%>%
  select(PFMA,YEAR,MONTH,MANAGEMENT,ESTIMATE,ESTIMATE,STANDARD_ERROR,PERCENT_STANDARD_ERROR)

e22<-merge(e22.prep,new.effort.area.merge,by="PFMA")%>%
  mutate(PFMA=factor(PFMA,levels=c("PFMA 11","PFMA 111","PFMA 12",
                                   "PFMA 13","PFMA 14","PFMA 15",
                                   "PFMA 16","PFMA 17","PFMA 18",
                                   "PFMA 19","PFMA 20","PFMA 21",
                                   "PFMA 22","PFMA 23","PFMA 24","PFMA 25","PFMA 26","PFMA 27",
                                   "PFMA 121","PFMA 123","PFMA 124","PFMA 125","PFMA 126","PFMA 127",
                                   "PFMA 28","PFMA 29")))


#x<-data.frame(PFMA=unique(e22.prep$PFMA))
#write.csv(x,"data/pfma to area groupings merge.csv")

years=seq(2012,2021,1)

e22$months=factor(e22$MONTH,levels=c("January","February","March","April","May","June","July","August","September","October","November","December"))

```

```{r 2024 data read and prep}
#| echo: false
#| message: false
#| warning: false

d22<-read_excel("data/catch and release data aug 2024 version.xlsx",sheet="2024 all months")

d22.long<-d22%>%mutate(encounters=kept+released)%>%
pivot_longer(c(5:8),names_to="disposition",values_to="estimate")%>%
  mutate(area2=factor(area))
```

```{r 2024 per SE data read}
#| echo: false
#| message: false
#| warning: false

perse<-read_excel("data/catch and release data aug 2024 version.xlsx",sheet="percent SE 2024 creel summary")%>%mutate(area2=factor(area))

```

```{r data prep}
#| echo: false
#| message: false
#| warning: false

gg.prep<-data%>%
  select(PFMA,CREEL_SUB_AREA,YEAR,MONTH,DISPOSITION,ADIPOSE_MARK,ESTIMATE,STANDARD_ERROR,PERCENT_STANDARD_ERROR,AREA=`Area Grouping`)

years=seq(2011,2021,1)

area.region.merge=data.frame(AREA=c("Alberni (23A,B)","Barkley (23C-Q)","Clayquot","Kyuquot",
                                    "Nitinat","Nootka/Esp","Quatsino",
                                    "NWVIoffshore(125,126,127)","SWVIoffshore(121,123,124)",
                                    "JDFeast(19cde,20cd)","JDFwest(Renfrew 20abe)",
                                    "nGST(13,14,15,16)","nJST(12)","sGST(17,18,19ab,28,29)","upperQCS(11,111)"),
                             region=c("WCVI","WCVI","WCVI","WCVI","WCVI","WCVI","WCVI","WCVI","WCVI",
                                      "JDF","JDF",
                                      "SOG/ECVI","SOG/ECVI","SOG/ECVI","SOG/ECVI"))

months=factor(c("May","June","July","August","September"),levels=c("May","June","July","August","September"))
month.merge<-data.frame(MONTH=c(5,6,7,8,9),Month=months)

gg.effort.prep<-merge(effort,month.merge,by="MONTH")

```

\newpage

# 2024 Creel Survey Summary

```{r 2024 creel survey summary}
#| label: fig-2024-creel-survey-summary
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Top: Total encounters of Chinook by area and area grouping for 2024 year to date and the 5-yr average. Bottom: Percentage change in total encounters (2024 YTD versus 5-yr average)."


d22.effort<-d22.long%>%filter(disposition=="effort")

e22.1<-ggplot(d22.effort,aes(x=area2,y=estimate,fill=year))+
  geom_col(position="dodge")+
  scale_fill_brewer(palette="Set1")+
  theme_bw()+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Effort",fill="Period")+
  theme(legend.position="bottom")

d22.effort.pchange<-d22.effort%>%pivot_wider(names_from="year",values_from="estimate")%>%
  mutate(pchange=(`2024`-`5-yr`)/`5-yr`*100)%>%
  select(area2,region,disposition,pchange)

e22.2<-ggplot(d22.effort.pchange,aes(x=area2,y=pchange))+
  geom_col(position="dodge")+
  #scale_fill_brewer(palette="Set1")+
  theme_bw()+
  geom_hline(yintercept=0)+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Percent Change in Effort (2024 versus 5-yr average)")
```

```{r 2024 creel survey summary output effort}
#| label: fig-2024-creel-survey-summary-output-effort
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Top: Total effort by area and area grouping for 2024 year to date and the 5-yr average. Bottom: Percentage change in total effort (2024 YTD versus 5-yr average)."

e22.1/e22.2

```

```{r 2024 creel survey summary kept}
#| label: fig-2024-creel-summary-kept
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: ")."
 
d22.kept<-d22.long%>%filter(disposition=="kept")

c22.1<-ggplot(d22.kept,aes(x=area2,y=estimate,fill=year))+
  geom_col(position="dodge")+
  scale_fill_brewer(palette="Set1")+
  theme_bw()+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Kept Chinook",fill="Period")+
  theme(legend.position="bottom")

d22.kept.pchange<-d22.kept%>%pivot_wider(names_from="year",values_from="estimate")%>%
  mutate(pchange=(`2024`-`5-yr`)/`5-yr`*100)%>%
  select(area2,region,disposition,pchange)

c22.2<-ggplot(d22.kept.pchange,aes(x=area2,y=pchange))+
  geom_col(position="dodge")+
  #scale_fill_brewer(palette="Set1")+
  theme_bw()+
  geom_hline(yintercept=0)+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Percent Change in Kept Catch\n(2024 versus 5-yr average)")
```

```{r 2024 creel survey summary output kept}
#| label: fig-2024-creel-summary-output-kept
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Top: Total kept catch of Chinook by area and area grouping for 2024 year to date and the 5-yr average. Bottom: Percentage change in total kept catch (2024 YTD versus 5-yr average)."

c22.1/c22.2

```

```{r 2024 creel survey summary release}
#| label: fig-2024-creel-summary-release
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: ")."

d22.rel<-d22.long%>%filter(disposition=="released")

r22.1<-ggplot(d22.rel,aes(x=area2,y=estimate,fill=year))+
  geom_col(position="dodge")+
  scale_fill_brewer(palette="Set1")+
  theme_bw()+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Released Chinook",fill="Period")+
  theme(legend.position="bottom")

d22.rel.pchange<-d22.rel%>%pivot_wider(names_from="year",values_from="estimate")%>%
  mutate(pchange=(`2024`-`5-yr`)/`5-yr`*100)%>%
  select(area2,region,disposition,pchange)

r22.2<-ggplot(d22.rel.pchange,aes(x=area2,y=pchange))+
  geom_col(position="dodge")+
  #scale_fill_brewer(palette="Set1")+
  theme_bw()+
  geom_hline(yintercept=0)+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Percent Change in Released Catch\n(2024 versus 5-yr average)")
```

```{r 2024 creel survey summary output released}
#| label: fig-2024-creel-summary-output-released
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Top: Total released catch of Chinook by area and area grouping for 2024 year to date and the 5-yr average. Bottom: Percentage change in total released catch (2024 YTD versus 5-yr average)."

r22.1/r22.2

```

```{r 2024 creel survey summary encounters}
#| label: fig-2024-creel-summary-encounters
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: ")."

d22.enc<-d22.long%>%filter(disposition=="encounters")

enc22.1<-ggplot(d22.enc,aes(x=area2,y=estimate,fill=year))+
  geom_col(position="dodge")+
  scale_fill_brewer(palette="Set1")+
  theme_bw()+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Total Encounters of Chinook",fill="Period")+
  theme(legend.position="bottom")

d22.enc.pchange<-d22.enc%>%pivot_wider(names_from="year",values_from="estimate")%>%
  mutate(pchange=(`2024`-`5-yr`)/`5-yr`*100)%>%
  select(area2,region,disposition,pchange)

enc22.2<-ggplot(d22.enc.pchange,aes(x=area2,y=pchange))+
  geom_col(position="dodge")+
  #scale_fill_brewer(palette="Set1")+
  theme_bw()+
  geom_hline(yintercept=0)+
  facet_grid(.~region,space="free",scale="free",switch="x",labeller = labeller(region = label_wrap_gen(10)))+
  labs(x="Area",y="Percent Change in Total Encounters\n(2024 versus 5-yr average)")
```

```{r 2024 creel survey summary output encounters}
#| label: fig-2024-creel-summary-output-encounters
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Top: Total encounters of Chinook by area and area grouping for 2024 year to date and the 5-yr average. Bottom: Percentage change in total encounters (2024 YTD versus 5-yr average)."

enc22.1/enc22.2
```

```{r 2024 creel survey summary per SE}
#| label: fig-2024-creel-summary-perSE
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Percent Standard Error for Chinook total effort, kept catch, and released catch estimates for 2024 year to date, by area and area grouping."

perse2<-perse%>%
  mutate(use=case_when(perse<=25~"relatively certain",
                       perse>=25&perse<=50~"use with caution",
                       perse>50~"high uncertainty"))

ggplot(perse2,aes(x=area2,y=perse,fill=factor(use,levels=c("relatively certain","use with caution","high uncertainty"))))+
         geom_col()+theme_bw()+
         facet_grid(disposition~region,space="free",switch="x",scale="free_x",labeller = labeller(region = label_wrap_gen(10)))+
  scale_fill_manual(values=c("dark green","orange","firebrick"))+
  labs(x="Area",y="Percent or Relative Standard Error",fill="")+
  theme(legend.position = "bottom")

ggsave("outputs/chinook percent standard error 2024 creel survey.png",dpi=600,height=6,width=7)

```

```{r 2024 creel survey summary stat areas}
#| label: fig-2024-creel-summary-stats
#| echo: false
#| message: false
#| warning: false

d22.long %>%
  ggplot(aes(x = area2, y = estimate, fill = month)) +
  geom_col()
```


# Effort Data Including 2022

-   data from DFO
-   2002-2022 by month by year by PFMA by management (ISBM versus AABM)
-   effort is total boat trips
-   unclear whether this is Chinook directed or total recreational effort (including other species etc.)

```{r total effort all sbc new data}
#| label: fig-total-effort-all-sbc-new
#| echo: false
#| message: false
#| warning: false
#| fig-height: 4
#| fig-width: 5
#| fig-cap: "Total effort in number of boat trips for all SBC areas."

e22.tot<-e22%>%group_by(YEAR)%>%summarise(total=sum(ESTIMATE,na.rm=TRUE))

ggplot(e22.tot,aes(x=YEAR,y=total))+
  geom_col()+
  theme_bw()+
  labs(x="Year",y="Total Effort SBC (boat trips)")
```

\newpage

```{r total effort all sbc by area by year}
#| label: fig-total-effort-all-sbc-area-yr
#| echo: false
#| message: false
#| warning: false
#| fig-height: 8.5
#| fig-width: 7
#| fig-cap: "Total effort in number of boat trips by area for all SBC areas."

e22.byarea<-e22%>%group_by(PFMA,YEAR,AREA)%>%summarise(effort=sum(ESTIMATE,na.rm=TRUE))%>%
  mutate(PFMA=factor(PFMA,levels=c("PFMA 11","PFMA 111","PFMA 12",
                                   "PFMA 13","PFMA 14","PFMA 15",
                                   "PFMA 16","PFMA 17","PFMA 18",
                                   "PFMA 19","PFMA 20","PFMA 21",
                                   "PFMA 22","PFMA 23","PFMA 24","PFMA 25","PFMA 26","PFMA 27",
                                   "PFMA 121","PFMA 123","PFMA 124","PFMA 125","PFMA 126","PFMA 127",
                                   "PFMA 28","PFMA 29")))

cols<-brewer.pal(7,"Set1")[c(1:5,7)]

ggplot(e22.byarea,aes(x=YEAR,y=effort,color=AREA))+
  geom_point()+
  geom_line()+
  theme_bw()+
  scale_color_manual(values=cols)+
  facet_wrap(~PFMA,scale="free_y",ncol=3)+
  theme(legend.position="bottom")+
  labs(x="Year",y="Effort (boat trips)",color="Area")

```

\newpage

```{r total effort all sbc by area by year only may-sept}
#| label: fig-total-effort-may-sept
#| echo: false
#| include: false
#| message: false
#| warning: false
#| fig-height: 8.5
#| fig-width: 7
#| fig-cap: "Total effort in number of boat trips by area for all SBC areas-ONLY MAY to SEPTEMBER."

e22.byareaonlysomemonths<-e22%>%filter(MONTH%in%c("May","June","July","August","September"))%>%
  group_by(PFMA,YEAR,AREA)%>%
  summarise(effort=sum(ESTIMATE,na.rm=TRUE))

cols<-brewer.pal(7,"Set1")[c(1:5,7)]

ggplot(e22.byarea,aes(x=YEAR,y=effort,color=AREA))+
  geom_point()+
  geom_line()+
  theme_bw()+
  scale_color_manual(values=cols)+
  facet_wrap(~PFMA,scale="free_y",ncol=3)+
  theme(legend.position="bottom")+
  labs(x="Year",y="Effort (boat trips)",color="Area")

```

\newpage

```{r total by disposition all sbc}
#| label: fig-total-disposition
#| echo: false
#| message: false
#| warning: false
#| fig-height: 3
#| fig-width: 5
#| fig-cap: "Total encounters (column height) by disposition (kept, released and released sub-legal) Chinook for all SBC areas."

gg.totalbydis<-gg.prep%>%
  group_by(YEAR,DISPOSITION)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))

c1<-ggplot(gg.totalbydis,aes(x=YEAR,y=total,fill=DISPOSITION))+
  geom_col()+
  scale_fill_brewer(palette="Set1")+
  scale_x_continuous(labels=as.character(years),breaks=years)+
  theme_bw()+
  labs(x="Year",y="Chinook",fill="")+
  geom_text(aes(label = total), position=position_stack(vjust=0.5), colour = "black",size=2)+
  theme(legend.position="bottom")
```

```{r total effort all sbc}
#| label: fig-total-effort-all-sbc
#| echo: false
#| message: false
#| warning: false
#| fig-height: 3
#| fig-width: 5
#| fig-cap: "Total Chinook directed effort by year and month for all SBC areas."

gg.efforttotal<-gg.effort.prep%>%filter(TARGET=="Chinook")%>%
  group_by(YEAR,Month)%>%
  summarise(total=sum(TRIPS,na.rm=TRUE))

e1<-ggplot(gg.efforttotal,aes(x=YEAR,y=total,fill=factor(Month)))+
  geom_col()+
  theme_bw()+
  scale_x_continuous(labels=as.character(seq(2014,2020,1)),breaks=seq(2014,2020,1))+
  scale_fill_brewer(palette="Set1")+
  labs(x="Year",y="Total Chinook Effort (trips)",fill="")+
  geom_text(aes(label = total), position=position_stack(vjust=0.5), colour = "black",size=3)+
  theme(legend.position="bottom")

#ggsave("figures/chinook total by disposition all SBC.png",height=5,width=6,dpi=600)
```

```{r total catch release effort all sbc}
#| label: fig-total-catch-release-all-sbc
#| echo: false
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 5
#| fig-cap: "Top: Total encounters (column height) by disposition (kept, released and released sub-legal) Chinook for all SBC areas. Bottom: Total Chinook directed effort by year and month for all SBC areas."

ggarrange(c1,e1,ncol=1)
```


\newpage

```{r total across disposition by area}
#| label: fig-total-across-disp-by-area
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Kept, released and released sub-legal Chinook by area."

gg.totalbydisbyarea<-gg.prep%>%
  group_by(AREA,YEAR,DISPOSITION)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))

ggplot(gg.totalbydisbyarea,aes(x=YEAR,y=total,fill=DISPOSITION))+
  geom_col()+
  scale_fill_brewer(palette="Set1")+
  scale_x_continuous(labels=as.character(years),breaks=years)+
  theme_bw()+
  facet_grid(AREA~DISPOSITION,scale="free_y",switch="y")+
  labs(x="Year",y="Chinook",fill="")+
 #geom_text(aes(label = total), position=position_stack(vjust=1), colour = "black",size=1,angle=90)+
  theme(legend.position="bottom",strip.text.y.left=element_text(angle=0,size=6))
```

```{r total across disposition by area for just some areas stacked}
#| label: fig-total-across-disp-stacked
#| echo: false
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 8
#| fig-cap: "Kept, released and released sub-legal Chinook by area for proposed MSF areas in SOG and JDF."

gg.totalbydisbyarea.msf<-gg.prep%>%
  filter(AREA%in%c("JDFeast(19cde,20cd)","JDFwest(Renfrew 20abe)","sGST(17,18,19ab,28,29)","nGST(13,14,15,16)"))%>%
  group_by(AREA,YEAR,DISPOSITION)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))

ggplot(gg.totalbydisbyarea.msf,aes(x=YEAR,y=total,fill=DISPOSITION))+
  geom_col()+
  scale_fill_brewer(palette="Set1")+
  scale_x_continuous(labels=as.character(years),breaks=years)+
  theme_bw()+
  facet_grid(AREA~.,scale="free_y",switch="y")+
  labs(x="Year",y="Number of Chinook",fill="")+
  geom_text(aes(label = total), position=position_stack(vjust=.5), colour = "black",size=2,angle=0)+
  theme(legend.position="bottom",strip.text.y.left=element_text(angle=0,size=6))

#ggsave("figures/chinook total by disposition by area msf filter.png",height=8,width=8,dpi=600)
```

\newpage

```{r effort by area month}
#| label: fig-effort-by-area-month
#| echo: false
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 8
#| fig-cap: "Chinook directed effort by area and month (2014-2020)"

gg.effortbyarea<-gg.effort.prep%>%
  group_by(area=LRG_AREA_NME,YEAR,Month)%>%
  summarise(effort=sum(TRIPS,na.rm=TRUE))

ggplot(gg.effortbyarea,aes(x=YEAR,y=effort,fill=Month))+
  #geom_line()+geom_point()+
  geom_col(position=position_dodge(preserve="single"))+
  scale_fill_brewer(palette="Set1")+
  theme_bw()+
  theme(legend.position = "bottom")+
  facet_wrap(~area,ncol=2)+
  labs(x="Year",y="Effort (trips)",color="")

```

\newpage

```{r total by disposition stacked by area}
#| label: fig-total-by-disp-stacked
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Kept, released and released sub-legal by area. Total column height is encounters."

gg.totalbydisbyarea<-gg.prep%>%
  group_by(AREA,YEAR,DISPOSITION)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))

#ggplot(gg.totalbydisbyarea,aes(x=YEAR,y=total,fill=DISPOSITION))+
ggplot(gg.totalbydisbyarea,aes(x=YEAR,y=total,color=DISPOSITION))+
  #geom_col()+
  geom_point()+geom_line()+
  #scale_fill_brewer(palette="Set1")+
  scale_color_brewer(palette="Set1")+
  scale_x_continuous(labels=as.character(years),breaks=years)+
  theme_bw()+
  facet_grid(AREA~DISPOSITION,scale="free_y",switch="y")+
  labs(x="Year",y="Chinook",fill="")+
#  geom_text(aes(label = total), position=position_stack(vjust=0.5), colour = "black",size=1)+
  theme(legend.position="bottom",strip.text.y.left=element_text(angle=0),axis.text.x = element_text(angle=90))

#ggsave("figures/chinook total by disposition by area totals.png",height=10,width=8,dpi=600)
#data%>%distinct(`Area Grouping`)

```

\newpage

```{r total encounters by area}
#| label: fig-total-encounters-by-area
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Total encounters by area 2011-2021. Lines fit by LOESS."
gg.encountersbyarea<-gg.prep%>%
  group_by(AREA,YEAR)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))

ggplot(gg.encountersbyarea,aes(x=YEAR,y=total))+
  geom_point()+geom_line()+
  #scale_fill_brewer(palette="Set1")+
  #scale_x_continuous(labels=as.character(years),breaks=years)+
  theme_bw()+
  geom_smooth()+
  facet_wrap(~AREA,scale="free_y",ncol=3)+
  labs(x="Year",y="Encounters")

#ggsave("figures/chinook total by disposition by area totals.png",height=10,width=8,dpi=600)
#data%>%distinct(`Area Grouping`)
```

\newpage

```{r total encounters by area base versus recent simple}
#| label: fig-encounters-by-area-base-versus-recent-simple
#| echo: false
#| message: false
#| warning: false
#| fig-height: 5
#| fig-width: 8
#| fig-cap: "Percent change in encounters between base period (2011-2019) and 2019-2021."

gg.encountersbaseversusrecentsimple<-gg.encountersbyarea%>%
  mutate(period=ifelse(YEAR<2019,"2011-2018","2019-2021"))%>%
  group_by(AREA,period)%>%
  summarise(u=mean(total))%>%
  pivot_wider(names_from="period",values_from="u")%>%
  mutate(pchange=`2019-2021`/`2011-2018`-1)%>%
  select(AREA,pchange)%>%
  merge(area.region.merge)%>%
  mutate(factor(region,levels=c("WCVI","SOG/ECVI","JDF")))

ggplot(gg.encountersbaseversusrecentsimple,aes(x=fct_reorder(AREA,region),y=pchange,fill=region))+
  geom_col()+
  theme_bw()+
  #facet_grid(. ~ region, space="free",scale="free")+ 
  labs(y="Percent Chnage in Encounters",x="Period")+
  scale_fill_brewer(palette="Set1")+
  theme(axis.text.x=element_text(angle=90))

```

\newpage

```{r total enucounters by area base versus recent}
#| label: fig-encounters-by-area-base-versus-recent
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Boxplot of total encounters in base period (2011-2018) versus recent years (2019-2021)."

gg.encountersbaseversusrecent<-gg.encountersbyarea%>%
  mutate(period=ifelse(YEAR<2019,"2011-2018","2019-2021"))

ggplot(gg.encountersbaseversusrecent,aes(x=period,y=total,fill=period))+
  geom_boxplot(notch=FALSE)+
  scale_fill_brewer(palette="Set1")+
  facet_wrap(~AREA,scale="free_y",ncol=3)+
  theme_bw()+
  labs(y="Encounters",x="Period")+
  guides(fill="none")
```

# Appendix A

```{r chinook directed effort}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Total Chinook directed effort based on data provided by DFO Stock Assessment (2014-2020)"

gg.effortbyarea<-gg.effort.prep%>%filter(TARGET=="Chinook")%>%
  group_by(LRG_AREA_NME,YEAR)%>%
  summarise(total=sum(TRIPS,na.rm=TRUE))

ggplot(gg.effortbyarea,aes(x=YEAR,y=total))+
  geom_point()+geom_line()+
  theme_bw()+
  facet_grid(LRG_AREA_NME~.,switch="y",scale="free_y")+
  theme(strip.text.y.left=element_text(angle=0,size=8))+
  labs(x="Year",y="Total Chinook Directed Effort")+
  expand_limits(y=0)

ggsave("figures/chinook directed effort by area.png",height=8,width=5,dpi=600)
```

