---
title: "2026 SBC Chinook Recreational Creel  Summary"
format: html
editor: visual
editor_options:
  chunk_output_type: console
---

## Data Sources

- 2010-2026 effort, catch and released numbers are based on DFO provided data.
- 2026 data is from the South Coast Assessment Creel Survey Summary Report and only contains data from June and July 2026.

## Notes

- sub-legal = smaller than the legal size limit
- encounters = total number of kept, released-legal, and released sub-legal Chinook

```{r load library}
#| echo: false
#| message: false
#| warning: false

knitr::opts_chunk$set(fig.pos = "H", out.extra = "")

options(scipen=1000)
library(data.table)
library(tidyverse)
library(RColorBrewer)
library(ggpubr)
library(stringr)
library(patchwork)
library(ggplot2)
library(dplyr)
require(readxl)
```

```{r data read}
#| echo: false
#| message: false
#| warning: false

# updated to 2026
data<-read_excel("data/2026_data/catch and release data july 2026 version.xlsx",sheet="chinook catch and release")

# updated to 2026
effort<-read_excel("data/2026_data/catch and release data july 2026 version.xlsx",sheet="effort by target species")
```

```{r data read new effort data - to 2026 July}
#| echo: false
#| message: false
#| warning: false 

# updated to 2026
new.effort.data<-fread("data/common/South_Coast_Recreational_effort_2012-2026.csv")

new.effort.area.merge<-fread(("data/common/pfma to area groupings merge.csv"))

e26.prep<-new.effort.data%>%
  select(PFMA,YEAR,MONTH,MANAGEMENT,ESTIMATE,ESTIMATE,STANDARD_ERROR,PERCENT_STANDARD_ERROR)

e26<-merge(e26.prep,new.effort.area.merge,by="PFMA")%>%
  mutate(PFMA=factor(PFMA,levels=c("PFMA 11","PFMA 111","PFMA 12",
                                   "PFMA 13","PFMA 14","PFMA 15",
                                   "PFMA 16","PFMA 17","PFMA 18",
                                   "PFMA 19","PFMA 20","PFMA 21",
                                   "PFMA 22","PFMA 23","PFMA 24",
                                   "PFMA 25","PFMA 26","PFMA 27",
                                   "PFMA 121","PFMA 123","PFMA 124",
                                   "PFMA 125","PFMA 126","PFMA 127",
                                   "PFMA 28","PFMA 29")))
years=seq(2012,2026,1)

e26$months=factor(e26$MONTH,levels=c("January","February","March","April","May","June","July","August","September","October","November","December"))
```

```{r 2026 data read and prep}
#| echo: false
#| message: false
#| warning: false

# chinook
d26<-read_excel("data/2026_data/catch and release data july 2026 version.xlsx",sheet="2026 creel survey chinook")

d26.long<-d26%>%mutate(encounters=kept+released) %>%
pivot_longer(c(5:7),names_to="disposition",values_to ="estimate") %>%
  mutate(area2=factor(area))

# coho
d26_co <-read_excel("data/2026_data/catch and release data july 2026 version.xlsx",sheet="2026 creel survey coho")

d26_co.long<-d26_co%>%mutate(encounters=kept+released) %>%
pivot_longer(c(5:7),names_to="disposition",values_to="estimate") %>%
  mutate(area2=factor(area))
```

```{r data prep}
#| echo: false
#| message: false
#| warning: false

gg.prep<-data%>%
  select(PFMA,CREEL_SUB_AREA,YEAR,MONTH,DISPOSITION,ADIPOSE_MARK,ESTIMATE,STANDARD_ERROR,PERCENT_STANDARD_ERROR,AREA=`Area Grouping`)

years=seq(2011,2026,1)

area.region.merge=data.frame(AREA=c("Alberni (23A,B)","Barkley (23C-Q)","Clayquot","Kyuquot",
                                    "Nitinat","Nootka/Esp","Quatsino",
                                    "NWVIoffshore(125,126,127)","SWVIoffshore(121,123,124)",
                                    "JDFeast(19cde,20cd)","JDFwest(Renfrew 20abe)",
                           "nGST(13,14,15,16)","nJST(12)","sGST(17,18,19ab,28,29)","upperQCS(11,111)"),
                             region=c("WCVI","WCVI","WCVI","WCVI","WCVI","WCVI","WCVI","WCVI","WCVI",
                                      "JDF","JDF",
                                      "SOG/ECVI","SOG/ECVI","SOG/ECVI","SOG/ECVI"))

months=factor(c("May","June","July","August","September"),levels=c("May","June","July","August","September"))
month.merge<-data.frame(MONTH=c(5,6,7,8,9),Month=months)

gg.effort.prep<-merge(effort,month.merge,by="MONTH")

```

# Annual Summary
```{r total by disposition all sbc}
#| label: fig-total-disposition
#| echo: false
#| message: false
#| warning: false
#| fig-height: 3
#| fig-width: 5
#| fig-cap: "Total encounters (column height) by disposition (kept, released and released sub-legal) Chinook for all SBC areas."

gg.totalbydis<-gg.prep%>%
  mutate(DISPOSITION = case_when(DISPOSITION %in% c("Released Legal", "Released Sub-legal")~ "Released", TRUE~DISPOSITION)) %>%
  group_by(YEAR,DISPOSITION)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))

c1<-ggplot(gg.totalbydis,aes(x=YEAR,y=total,fill=DISPOSITION))+
  geom_col()+
  scale_fill_viridis_d()+
  scale_x_continuous(labels=as.character(years),breaks = years)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(x="Year",y="Chinook",fill="")+
  geom_text(aes(label = total), position=position_stack(vjust=0.5), colour = "black",size=2)+
  theme(legend.position="bottom")
```

```{r total effort all sbc}
#| label: fig-total-effort-all-sbc
#| echo: false
#| message: false
#| warning: false
#| fig-height: 3
#| fig-width: 5
#| fig-cap: "Total Chinook directed effort by year and month for all SBC areas."

gg.efforttotal<-gg.effort.prep%>%filter(TARGET=="Chinook")%>%
  group_by(YEAR,Month)%>%
  summarise(total=sum(TRIPS,na.rm=TRUE))

e1<-ggplot(gg.efforttotal,aes(x=YEAR,y=total,fill=factor(Month)))+
  geom_col()+
  theme_bw()+
  scale_x_continuous(labels=as.character(seq(2014,2026,1)),breaks=seq(2014,2026,1)) +
  scale_fill_viridis_d()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(x="Year",y="Total Chinook Effort (trips)",fill="")+
  geom_text(aes(label = total), position=position_stack(vjust=0.5), colour = "black",size=3)+
  theme(legend.position="bottom")

```

```{r total catch release effort all sbc}
#| label: fig-total-catch-release-all-sbc
#| echo: false
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 5
#| fig-cap: "Top: Total encounters (column height) by disposition (kept, released and released sub-legal) Chinook for all SBC areas. Bottom: Total Chinook directed effort by year and month for all SBC areas."

ggarrange(c1,e1,ncol=1)
```

```{r total across disposition by area for just some areas stacked}
#| label: fig-total-across-disp-stacked
#| echo: false
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 8
#| fig-cap: "Kept, released and released sub-legal Chinook by PFMA."

gg.totalbydisbyarea.msf<-gg.prep%>%
  mutate(DISPOSITION = case_when(DISPOSITION %in% c("Released Legal", "Released Sub-legal")~ "Released", TRUE~DISPOSITION)) %>%
  filter(AREA%in%c("JDFeast(19cde,20cd)","JDFwest(Renfrew 20abe)","sGST(17,18,19ab,28,29)","nGST(13,14,15,16)"))%>%
  group_by(PFMA,YEAR,DISPOSITION)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))

ggplot(gg.totalbydisbyarea.msf,aes(x=YEAR,y=total,fill=DISPOSITION))+
  geom_col()+
  scale_fill_viridis_d()+
  scale_x_continuous(labels=as.character(years),breaks=years)+
  theme_bw()+
  facet_grid(PFMA~.,scale="free_y",switch="y")+
  labs(x="Year",y="Number of Chinook",fill="")+
  geom_text(aes(label = total), position=position_stack(vjust=.5), colour = "black",size=2,angle=0)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(legend.position="bottom",strip.text.y.left=element_text(angle=0,size=6))

#ggsave("figures/chinook total by disposition by area msf filter.png",height=8,width=8,dpi=600)
```

\newpage

```{r effort by area month}
#| label: fig-effort-by-area-month
#| echo: false
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 8
#| fig-cap: "Chinook directed effort by area and month (2014-2026)"

# gg.effortbyarea<-gg.effort.prep%>%
#   group_by(area=LRG_AREA_NME,YEAR,Month)%>%
#   summarise(effort=sum(TRIPS,na.rm=TRUE))
# 
# ggplot(gg.effortbyarea,aes(x=YEAR,y=effort,fill=Month))+
#   geom_col(position=position_dodge(preserve="single"))+
#   scale_fill_viridis_d()+
#   theme_bw()+
#   scale_x_continuous(labels=as.character(seq(2014,2026,1)),breaks=seq(2014,2026,1))+
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))+
#   theme(legend.position = "bottom")+
#   facet_wrap(~area,ncol=2)+
#   labs(x="Year",y="Effort (trips)",color="")

```

```{r total by disposition stacked by area}
#| label: fig-total-by-disp-stacked
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Kept, released and released sub-legal by area. Total column height is encounters."

gg.totalbydisbyarea<-gg.prep%>%
  mutate(DISPOSITION = case_when(DISPOSITION %in% c("Released Legal", "Released Sub-legal")~ "Released", TRUE~DISPOSITION)) %>%
  group_by(PFMA,YEAR,DISPOSITION)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))

ggplot(gg.totalbydisbyarea,aes(x=YEAR,y=total,color=DISPOSITION))+
  geom_point()+geom_line()+
  scale_x_continuous(labels=as.character(years),breaks=years)+
  theme_bw()+
  facet_grid(PFMA~DISPOSITION,scale="free_y",switch="y")+
  labs(x="Year",y="Chinook",fill="")+
  theme(legend.position="none",strip.text.y.left=element_text(angle=0),
        axis.text.x = element_text(angle=60, hjust = 1))

```

```{r total encounters by area}
#| label: fig-total-encounters-by-area
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Total encounters by area 2011-2026. Lines fit by LOESS."
gg.encountersbyarea<-gg.prep%>%
  group_by(PFMA,YEAR)%>%
  summarise(total=sum(ESTIMATE,na.rm=TRUE))

ggplot(gg.encountersbyarea,aes(x=YEAR,y=total))+
  geom_point()+geom_line()+
  theme_bw()+
  geom_smooth()+
  scale_x_continuous(labels=as.character(years),breaks=years)+
  facet_wrap(~PFMA,scale="free_y",ncol=3)+
  labs(x="Year",y="Encounters")+
  theme(axis.text.x = element_text(angle=60, hjust = 1))
```

```{r total encounters by area base versus recent simple}
#| label: fig-encounters-by-area-base-versus-recent-simple
#| echo: false
#| message: false
#| warning: false
#| fig-height: 5
#| fig-width: 8
#| fig-cap: "Percent change in encounters between base period (2011-2021) and 2022-2026."

gg.encountersbaseversusrecentsimple<-gg.encountersbyarea%>%
  mutate(period=ifelse(YEAR<2022,"2011-2021","2022-2026"))%>%
  group_by(PFMA,period)%>%
  summarise(u=mean(total))%>%
  pivot_wider(names_from="period",values_from="u")%>%
  mutate(pchange=`2022-2026`/`2011-2021`-1)%>%
  select(PFMA,pchange)%>%
  merge(area.region.merge)%>%
  mutate(factor(region,levels=c("WCVI","SOG/ECVI","JDF")))

ggplot(gg.encountersbaseversusrecentsimple,aes(x=fct_reorder(PFMA,region),y=pchange,fill=region))+
  geom_col()+
  theme_bw()+
  labs(y="Percent Change in Encounters",x="Period")+
  scale_fill_viridis_d()+
  theme(axis.text.x=element_text(angle=90))

```

```{r total encounters by area base versus recent}
#| label: fig-encounters-by-area-base-versus-recent
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Boxplot of total encounters in base period (2011-2021) versus recent years (2022-2026)."

gg.encountersbaseversusrecent<-gg.encountersbyarea%>%
  mutate(period=ifelse(YEAR<2022,"2011-2021","2022-2026"))

ggplot(gg.encountersbaseversusrecent,aes(x=period,y=total,fill=period))+
  geom_boxplot(notch=FALSE)+
  scale_fill_viridis_d()+
  facet_wrap(~PFMA,scale="free_y",ncol=3)+
  theme_bw()+
  labs(y="Encounters",x="Period")+
  guides(fill="none")
```

# 2026 Summary

## Chinook effort, catch and release

```{r chinook effort area * month}
#| label: fig-cn-effort
#| echo: false
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 5
#| fig-cap: "Annual Chinook effort by month and Area, 2002-2026."

e26 <- e26 %>%
  filter(months %in% c("May", "June", "July", "August", "September")) %>%
  mutate(months = factor(months, levels = c("May", "June", "July", "August", "September"))) %>%
  group_by(YEAR, `AREA`, months) %>%
  summarise(estimate = sum(ESTIMATE, na.rm = TRUE),.groups = "drop")

ggplot(e26, aes(x = YEAR, y = estimate, fill = months)) +
  geom_col() +
  scale_fill_viridis_d()+
  facet_wrap(~`AREA`) +
  labs(x = "Year",
    y = "Estimated effort (boat trips)",
    title = "Chinook salmon – Effort") +
  theme_bw()+
  scale_x_continuous(breaks = seq(2002, 2026, by = 4))+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  theme(legend.position = "bottom")

```

```{r chinook kept area * month}
#| label: fig-cn-kept
#| echo: false
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 5
#| fig-cap: "Annual Chinook kept catch by month and Area, 2011-2026."

chinook_kept <- data %>%
  filter(SPECIES == "CHINOOK SALMON", DISPOSITION == "Kept") %>%
  filter(MONTH %in% c("June", "July")) %>%
  mutate(MONTH = factor(MONTH, levels = c("May", "June", "July", "August", "September"))) %>%
  group_by(YEAR, PFMA, MONTH) %>%
  summarise(estimate = sum(ESTIMATE, na.rm = TRUE),.groups = "drop")

# ggplot(chinook_kept, aes(x = YEAR, y = estimate)) +
#   geom_col() +
#   scale_fill_viridis_d()+
#   facet_grid(`Area Grouping`~MONTH) +
#   labs(x = "Year",
#     y = "Estimated catch",
#     title = "Chinook salmon – Kept catch") +
#   theme_bw()+
#   theme(legend.position = "bottom")

ggplot(chinook_kept, aes(x = YEAR, y = estimate, fill = YEAR == 2026)) +
  geom_col() +
  scale_fill_manual(values = c(`FALSE` = "grey50", `TRUE` = "#31688E"),
    labels = c(`FALSE` = "Previous years", `TRUE` = "2026"), name = NULL) +
  scale_x_continuous(labels=as.character(seq(2010,2026,2)),
                     breaks=seq(2010,2026,2))+
  scale_y_continuous(labels = as.character(seq(0, 12500, 5000)), breaks = seq(0,12500,5000))+
  facet_grid(PFMA ~ MONTH) +
  labs(x = "Year", y = "Estimated catch",
    title = "Chinook salmon – Kept catch") +
  theme_bw() +
  theme(legend.position = "none")+
  theme(strip.text.y = element_text(angle = 0))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r chinook released area * month}
#| label: fig-cn-rel
#| echo: false
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 5
#| fig-cap: "Annual Chinook released catch by month and Area, 2011-2026."

chinook_rel <- data %>%
  mutate(DISPOSITION = case_when(DISPOSITION %in% c("Released Legal", "Released Sub-legal")~ "Released",
                                   TRUE~DISPOSITION)) %>%
  filter(SPECIES == "CHINOOK SALMON", DISPOSITION == "Released") %>%
  filter(MONTH %in% c("June", "July")) %>%
  mutate(MONTH = factor(MONTH, levels = c("May", "June", "July", "August", "September"))) %>%
  group_by(YEAR, PFMA, MONTH) %>%
  summarise(estimate = sum(ESTIMATE, na.rm = TRUE),.groups = "drop")

# ggplot(chinook_kept, aes(x = YEAR, y = estimate, fill = MONTH)) +
#   geom_col() +
#   scale_fill_viridis_d()+
#   facet_wrap(~PFMA) +
#   labs(x = "Year",
#     y = "Estimated catch",
#     title = "Chinook salmon – Released catch") +
#   theme_bw()+
#   theme(legend.position = "bottom")

ggplot(chinook_rel, aes(x = YEAR, y = estimate, fill = YEAR == 2026)) +
  geom_col() +
  scale_fill_manual(
    values = c(`FALSE` = "grey50", `TRUE` = "#31688E"),
    labels = c(`FALSE` = "Previous years", `TRUE` = "2026"),
    name = NULL) +
  scale_x_continuous(labels=as.character(seq(2010,2026,2)),
                     breaks=seq(2010,2026,2))+
  facet_grid(PFMA ~ MONTH) +
  labs(x = "Year",
    y = "Estimated catch (released)",
    title = "Chinook salmon – Released catch") +
  theme_bw() +
  theme(legend.position = "none")+
  theme(strip.text.y = element_text(angle = 0))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r encounters Chinook}
#| label: fig-cn-encounters
#| echo: false
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 5
#| fig-cap: "Annual Chinook total encounters by month and Area, 2011-2026."
chinook_en <- data %>%
  filter(MONTH %in% c("June", "July")) %>%
  mutate(MONTH = factor(MONTH, levels = c("May", "June", "July", "August", "September"))) %>%
  group_by(YEAR, PFMA, MONTH) %>%
  summarise(encounters = sum(ESTIMATE, na.rm = TRUE), .groups = "drop")


ggplot(chinook_en, aes(x = YEAR, y = encounters, fill = YEAR == 2026)) +
  geom_col() +
  scale_fill_manual(
    values = c(`FALSE` = "grey50", `TRUE` = "#31688E"),
    labels = c(`FALSE` = "Previous years", `TRUE` = "2026"),
    name = NULL) +
  scale_x_continuous(labels=as.character(seq(2010,2026,2)),
                     breaks=seq(2010,2026,2))+
  scale_y_continuous(labels = as.character(seq(0, 30000, 15000)),
                     breaks = seq(0, 30000, 15000))+
  facet_grid(PFMA ~ MONTH) +
  labs(x = "Year",
    y = "Estimated encounters",
    title = "Chinook salmon – Total encounters") +
  theme_bw() +
  theme(legend.position = "none")+
  theme(strip.text.y = element_text(angle = 0))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r 2026 summary chinook}
#| echo: false
#| message: false
#| warning: false
d26_26 <- d26.long %>%
  filter(year == "2026") %>%
  mutate(month = factor(month, levels = c("June", "July", "August", "September")),
         Area = factor(area))

cnplot_26 <- ggplot(d26_26, aes(x = region, y = estimate, fill = Area)) +
  geom_col() +
  theme_bw() +
  facet_grid(disposition~month) +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=0.5)) +
  ggtitle("2026 Chinook effort and catch estimates by month and region") +
  labs(x = "Region", y = "Estimate") +
    scale_fill_viridis_d()

```

```{r 2026 summary coho}
#| echo: false
#| message: false
#| warning: false
d26_coho <- d26_co.long %>%
  filter(year == "2026") %>%
  mutate(month = factor(month, levels = c("June", "July", "August", "September")),
         Area = factor(area))

coplot_26 <- ggplot(d26_coho, aes(x = region, y = estimate, fill = Area)) +
  geom_col() +
  theme_bw() +
  facet_grid(disposition~month) +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=0.5)) +
  ggtitle("2026 Coho effort and catch estimates by month and region") +
  labs(x = "Region", y = "Estimate") +
    scale_fill_viridis_d()

```

## 2026 vs 5yr averages - Chinook

```{r 2026 creel survey summary output cn}
#| label: fig-2026-creel-survey-summary-output-effort
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Total effort, kept catch, and released catch by area and month for 2026 year to date for Chinook."

cnplot_26
```

## 2026 vs 5yr averages - Coho

```{r 2026 creel survey summary output co}
#| label: fig-2026-creel-survey-summary-effort-co
#| echo: false
#| message: false
#| warning: false
#| fig-height: 9
#| fig-width: 8
#| fig-cap: "Total effort, kept catch, and released catch by area and month for 2026 year to date for Coho."

coplot_26
```

```{r coho effort area * month}


```

```{r coho kept area * month}


```

```{r coho released area * month}


```



# Appendix A

```{r chinook directed effort}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Total Chinook directed effort based on data provided by DFO Stock Assessment (2014-2026)"

gg.effortbyarea<-gg.effort.prep%>%filter(TARGET=="Chinook")%>%
  group_by(LRG_AREA_NME,YEAR)%>%
  summarise(total=sum(TRIPS,na.rm=TRUE))

ggplot(gg.effortbyarea,aes(x=YEAR,y=total))+
  geom_point()+geom_line()+
  theme_bw()+
  facet_grid(LRG_AREA_NME~.,switch="y",scale="free_y")+
  theme(strip.text.y.left=element_text(angle=0,size=8))+
  labs(x="Year",y="Total Chinook Directed Effort")+
  expand_limits(y=0)
```

